---
title: "Final project notebook: earnings calls and CRSP daily returns"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen = 999)
set.seed(42)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(tibble)
library(stringr)
library(lubridate)
library(zoo)
library(slider)
library(ggplot2)
library(reshape2)
library(lightgbm)
library(xgboost)
```

This notebook builds a **daily security–date panel**: CRSP daily fields merged with **prior-quarter** transcript-based metrics, then models **next-day `dlyretx`** (with `next_day_return` from `dlyret` kept for diagnostics) using **LightGBM**, **XGBoost**, optional **Random Forest**, and **walk-forward time-series CV** for hyperparameter tuning.

## How to run

Run chunks **top to bottom** in order. The first code chunk loads R packages; paths point to `Data/Raw Data/` under this project (adjust if your layout differs). Later sections mirror the Python notebook’s **GPU** notes where relevant; this R port runs **LightGBM / XGBoost on CPU** by default.

## Main outputs

- Cleaned transcript universe aligned with CRSP tickers
- `merged_daily_df` / `model_df` with transcript **missingness flags**, **z-scored** transcript metrics and dynamics (train-fitted scaler), **`dlyret` / `dlyretx` lags**, rolling stats, **OHLC microstructure**, optional **cross-sectional** features, **`FE_STAGE`** ablation switch, and encoded categoricals (`ticker_enc`, `siccd_enc`, `naics_enc`)
- Time-based train / validation / test split and model comparison metrics

## 1. Data loading

- Load CRSP daily (`crsp_v2_2014_2024.csv`) and the wide transcript metrics file.
- Use `glimpse()` to confirm dtypes, row counts, and column names before cleaning.


```{r cell_load_csvs}
# --- Load raw CRSP and transcript CSVs; print schema ---
crsp_df <- read_csv(
  "H:/My Drive/Project/FA Project/Data/Raw Data/crsp_v2_2014_2024.csv",
  show_col_types = FALSE
)
transcripts_df <- read_csv(
  "H:/My Drive/Project/FA Project/Data/Raw Data/Transcripts Machine Readable Data.csv",
  show_col_types = FALSE
)
glimpse(crsp_df)
glimpse(transcripts_df)
```

## 2. Ticker coverage (transcripts vs CRSP)

Identify symbols that appear in **transcripts** but not in **CRSP**, and decide how to restrict the transcript universe so merges are reliable.


### Tickers in transcripts missing from CRSP

List symbols present in the transcript file but **not** in CRSP `ticker`, before we restrict the transcript universe.

```{r cell_ticker_compare}
# --- Compare unique tickers: transcripts vs CRSP ---
crsp_tickers <- unique(crsp_df$ticker)
transcripts_tickers <- unique(transcripts_df$Symbol)
cat(length(unique(transcripts_df$Symbol)), "---", length(unique(crsp_df$ticker)), "\n")
for (ticker in transcripts_tickers) {
  if (!ticker %in% crsp_tickers) {
    co_name <- unique(transcripts_df$`Company Name`[transcripts_df$Symbol == ticker])
    cat(ticker, paste0("['", paste(co_name, collapse = "', '"), "']"), "\n")
  }
}
```

### Tickers dropped from transcripts

The previous chunk lists companies present in **transcripts** but not in **CRSP** (e.g. alternate share classes). Those tickers are removed from the transcript table so later merges use a consistent symbol set.

## 3. Exploratory missingness

Quick **null counts** on the raw CRSP and transcript tables after the initial universe decision.


```{r cell_crsp_nulls}
# --- CRSP: null counts per column ---
colSums(is.na(crsp_df))
```

```{r cell_transcripts_nulls}
# --- Transcripts: null counts per column ---
colSums(is.na(transcripts_df))
```

## 4. Transcript filtering and quarterly missingness

Keep transcript rows in the CRSP ticker set, analyze **quarterly (CQ*)** missing values by ticker, optionally drop the worst tickers, and summarize missing rates.


```{r cell_filter_transcripts}
# 1) Filter transcript rows for symbols absent from CRSP
crsp_tickers_set <- unique(na.omit(crsp_df$ticker))
transcript_tickers_set <- unique(na.omit(transcripts_df$Symbol))
missing_symbols <- sort(setdiff(transcript_tickers_set, crsp_tickers_set))

cat("Symbols in transcripts but not in CRSP:", paste0("['", paste(missing_symbols, collapse = "', '"), "']"), "\n")
cat("Count:", length(missing_symbols), "\n")

transcripts_filtered_df <- transcripts_df %>%
  filter(!Symbol %in% missing_symbols)

# Drop unnamed columns if present
unnamed_cols <- grep("^Unnamed", colnames(transcripts_filtered_df), value = TRUE)
if (length(unnamed_cols) > 0) {
  transcripts_filtered_df <- transcripts_filtered_df %>% select(-all_of(unnamed_cols))
}

cat("Unique symbols before:", length(unique(transcripts_df$Symbol)), "\n")
cat("Unique symbols after:", length(unique(transcripts_filtered_df$Symbol)), "\n")
cat("Rows before:", nrow(transcripts_df), "| Rows after:", nrow(transcripts_filtered_df), "\n")
```

```{r cell_ohlc_missing}
# 2) Identify which tickers have missing values in OHLC columns
ohlc_cols <- c("dlyclose", "dlyhigh", "dlyopen", "dlylow")
missing_ohlc_mask <- rowSums(is.na(crsp_df[, ohlc_cols])) > 0

missing_ohlc_ticker_counts <- sort(table(crsp_df$ticker[missing_ohlc_mask]), decreasing = TRUE)
cat("Ticker-level missing OHLC counts:\n")
print(missing_ohlc_ticker_counts)

missing_ohlc_details <- crsp_df[missing_ohlc_mask, c("permno", "ticker", "dlycaldt", ohlc_cols)] %>%
  arrange(ticker, dlycaldt)

cat("\nSample missing OHLC rows:\n")
print(head(missing_ohlc_details, 20))
```

```{r cell_missing_per_quarter_plot}
# 3) Plot missing count by quarter in transcripts (filtered universe)
quarter_cols <- grep("^CQ[1-4]\\d{4}$", colnames(transcripts_filtered_df), value = TRUE)

quarter_sort_key <- function(col_name) {
  q <- as.integer(substr(col_name, 3, 3))
  y <- as.integer(substr(col_name, 4, nchar(col_name)))
  y * 10 + q
}
quarter_cols <- quarter_cols[order(sapply(quarter_cols, quarter_sort_key))]
missing_per_quarter <- colSums(is.na(transcripts_filtered_df[, quarter_cols]))

missing_per_quarter_df <- tibble(
  quarter = factor(quarter_cols, levels = quarter_cols),
  missing_count = as.integer(missing_per_quarter)
)

ggplot(missing_per_quarter_df, aes(x = quarter, y = missing_count)) +
  geom_col(fill = "steelblue") +
  labs(title = "Missing Transcript Count by Quarter (Filtered Symbols)",
       x = "Quarter", y = "Missing count") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

```{r cell_ticker_missing_summary}
# 4) Ticker with highest transcript missing values vs total quarterly cells
symbol_rows <- transcripts_filtered_df %>%
  group_by(Symbol) %>%
  summarise(n_rows = n(), .groups = "drop")

missing_cells_df <- transcripts_filtered_df %>%
  group_by(Symbol) %>%
  summarise(missing_cells = sum(is.na(across(all_of(quarter_cols)))), .groups = "drop")

ticker_missing_summary <- symbol_rows %>%
  inner_join(missing_cells_df, by = "Symbol") %>%
  mutate(
    total_quarterly_cells = n_rows * length(quarter_cols),
    missing_rate = missing_cells / total_quarterly_cells
  ) %>%
  arrange(desc(missing_cells))

cat("Top 10 tickers by missing quarterly transcript values:\n")
print(head(ticker_missing_summary, 20))

ggplot(ticker_missing_summary, aes(x = seq_along(missing_rate), y = missing_rate)) +
  geom_line() +
  labs(x = "ticker rank", y = "missing_rate")
```

### Tickers with heavy quarterly missingness

Some tickers had **entirely** or **mostly** missing `CQ*` transcript fields. The following chunks **drop rows** for the top four tickers by missing quarterly cells, then re-check missing counts.

```{r cell_drop_top4}
# 5) Remove rows for the top 4 tickers with the most missing quarterly transcript values
top4_missing_tickers <- head(ticker_missing_summary$Symbol, 4)
cat("Dropping rows for top 4 missing tickers:", paste0("['", paste(top4_missing_tickers, collapse = "', '"), "']"), "\n")

rows_before <- nrow(transcripts_filtered_df)
syms_before <- length(unique(transcripts_filtered_df$Symbol))

transcripts_filtered_df <- transcripts_filtered_df %>%
  filter(!Symbol %in% top4_missing_tickers)

cat("Rows before drop:", rows_before, "| after:", nrow(transcripts_filtered_df), "\n")
cat("Unique symbols before/after:", syms_before, "|", length(unique(transcripts_filtered_df$Symbol)), "\n")
```

```{r cell_new_missing_per_quarter}
# 6) New missing values after deleting top 4 missing tickers
new_missing_per_quarter <- colSums(is.na(transcripts_filtered_df[, quarter_cols]))
cat("New missing values per quarter after deletion:\n")
print(new_missing_per_quarter)
cat("\nTotal missing quarterly values after deletion:", sum(new_missing_per_quarter), "\n")

new_missing_df <- tibble(
  quarter = factor(quarter_cols, levels = quarter_cols),
  missing_count = as.integer(new_missing_per_quarter)
)

ggplot(new_missing_df, aes(x = quarter, y = missing_count)) +
  geom_col(fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

### After dropping high-missing transcript rows

Quarterly missing counts and a bar chart summarize remaining gaps after removing the worst tickers by missing `CQ*` cells.

```{r cell_quarter_missing_rate}
# 7) Average and median missing rate across quarters in transcripts_df
quarter_cols_all <- grep("^CQ", colnames(transcripts_df), value = TRUE)

quarter_missing_rate <- colMeans(is.na(transcripts_df[, quarter_cols_all])) * 100

cat("Missing rate (%) of each quarter:\n")
print(quarter_missing_rate)

avg_missing_rate <- mean(quarter_missing_rate)
median_missing_rate <- median(quarter_missing_rate)

cat(sprintf("\nAverage missing rate across quarters: %.2f%%\n", avg_missing_rate))
cat(sprintf("Median missing rate across quarters: %.2f%%\n", median_missing_rate))
```

## 5. CRSP universe alignment and column prep

Drop tickers removed from transcripts, inspect CRSP uniqueness, **drop unused columns**, and parse **`dlycaldt`** as datetimes for time-series operations.


```{r cell_remove_from_crsp}
# 8) Remove from CRSP the same tickers removed from transcripts
removed_from_transcripts <- c()
if (exists("missing_symbols")) removed_from_transcripts <- c(removed_from_transcripts, missing_symbols)
if (exists("top4_missing_tickers")) removed_from_transcripts <- c(removed_from_transcripts, top4_missing_tickers)
removed_from_transcripts <- sort(unique(removed_from_transcripts))
cat("Tickers removed from transcripts_df:", paste0("['", paste(removed_from_transcripts, collapse = "', '"), "']"), "\n")

crsp_rows_before <- nrow(crsp_df)
crsp_tickers_before <- length(unique(crsp_df$ticker))

crsp_df <- crsp_df %>% filter(!ticker %in% removed_from_transcripts)

cat("CRSP rows before/after:", crsp_rows_before, "|", nrow(crsp_df), "\n")
cat("CRSP unique tickers before/after:", crsp_tickers_before, "|", length(unique(crsp_df$ticker)), "\n")
```

```{r cell_crsp_head}
# --- CRSP sample rows after alignment ---
head(crsp_df)
```

```{r cell_crsp_nunique}
# --- CRSP cardinality of key columns ---
sapply(crsp_df, function(col) length(unique(col)))
```

```{r cell_crsp_drop_cols}
# 10) Drop selected identifier/metadata columns from crsp_df
cols_to_drop <- c(
  "permno", "permco", "cusip", "issuernm",
  "securitytype", "sharetype", "usincflg", "shradrflg"
)

existing_cols_to_drop <- intersect(cols_to_drop, colnames(crsp_df))

crsp_df <- crsp_df %>% select(-all_of(existing_cols_to_drop))

cat("Dropped columns from crsp_df:", paste0("['", paste(existing_cols_to_drop, collapse = "', '"), "']"), "\n")
cat("New crsp_df shape:", paste(nrow(crsp_df), ncol(crsp_df), sep = ", "), "\n")
```

```{r cell_crsp_parse_date}
# Convert dlycaldt to Date dtype
crsp_df <- crsp_df %>% mutate(dlycaldt = ymd(dlycaldt))
```

```{r cell_transcripts_shape}
# --- Transcript panel shape after filtering ---
dim(transcripts_filtered_df)
```

## 6. Transcript reshape (ticker × quarter)

Pivot wide transcript data into **one row per (ticker, fiscal quarter)** with the modeling metrics as columns.


```{r cell_reshape_transcripts}
# 11) Reshape transcripts into one row per ticker-quarter with 4 metric columns
source_transcripts_df <- if (exists("transcripts_filtered_df")) transcripts_filtered_df else transcripts_df

source_transcripts_df <- source_transcripts_df %>%
  rename(ticker = Symbol) %>%
  mutate(ticker = str_trim(as.character(ticker)))

metric_col_candidates <- colnames(source_transcripts_df)[tolower(str_trim(colnames(source_transcripts_df))) == "metrics"]
metric_col <- metric_col_candidates[1]
colnames(source_transcripts_df)[colnames(source_transcripts_df) == metric_col] <- "metric_name"
source_transcripts_df$metric_name <- str_trim(as.character(source_transcripts_df$metric_name))

metric_map <- c(
  "net positivity" = "net_positivity",
  "language complexity" = "language_complexity",
  "numeric transeprency" = "numeric_transparency",
  "numeric transparency" = "numeric_transparency",
  "analyst selectivity ratio" = "analyst_selectivity_ratio"
)

source_transcripts_df <- source_transcripts_df %>%
  mutate(metric_key = str_replace_all(tolower(metric_name), "\\s+", " "),
         metric_key = unname(metric_map[metric_key]))

quarter_cols <- grep("^CQ", colnames(source_transcripts_df), value = TRUE)
quarter_cols <- quarter_cols[order(sapply(quarter_cols, function(c) {
  as.integer(substr(c, 4, nchar(c))) * 10 + as.integer(substr(c, 3, 3))
}))]

id_cols <- c("ticker", "metric_key")

transcripts_long_df <- source_transcripts_df %>%
  select(all_of(c(id_cols, quarter_cols))) %>%
  pivot_longer(cols = all_of(quarter_cols),
               names_to = "quarter_code",
               values_to = "metric_value")

quarter_parts <- str_match(transcripts_long_df$quarter_code, "^CQ([1-4])(\\d{4})$")
transcripts_long_df$quarter_num <- as.integer(quarter_parts[, 2])
transcripts_long_df$year <- as.integer(quarter_parts[, 3])
transcripts_long_df <- transcripts_long_df %>%
  filter(!is.na(quarter_num) & !is.na(year))
transcripts_long_df$quarter_key <- as.yearqtr(paste0(transcripts_long_df$year, " Q", transcripts_long_df$quarter_num))

transcripts_metric_df <- transcripts_long_df %>%
  select(ticker, quarter_key, metric_key, metric_value) %>%
  pivot_wider(names_from = metric_key, values_from = metric_value)

metric_feature_cols <- c(
  "net_positivity",
  "language_complexity",
  "numeric_transparency",
  "analyst_selectivity_ratio"
)
for (col in metric_feature_cols) {
  if (!col %in% colnames(transcripts_metric_df)) {
    transcripts_metric_df[[col]] <- NA
  }
}

transcripts_metric_df <- transcripts_metric_df %>%
  select(ticker, quarter_key, all_of(metric_feature_cols))

cat("transcripts_long_df shape:", paste(nrow(transcripts_long_df), ncol(transcripts_long_df), sep = ", "), "\n")
cat("transcripts_metric_df shape:", paste(nrow(transcripts_metric_df), ncol(transcripts_metric_df), sep = ", "), "\n")
head(transcripts_metric_df)
```

```{r cell_transcripts_metric_preview}
# --- Preview ticker-quarter metric table ---
head(transcripts_metric_df, 50)
```

## 7. Daily panel merge and target

Join CRSP daily rows to the **prior completed quarter** transcript metrics, **deduplicate** `(ticker, date)` on CRSP and again after merge, build **`next_day_dlyretx`** (primary modeling target) and **`next_day_return`** (diagnostic, from `dlyret`), audit/drop rare non-last target NaNs, validate metric coverage.


```{r cell_build_merged}
# 12) Build ML panel — merge CRSP daily with prior-quarter transcript metrics
crsp_merge_df <- crsp_df %>%
  mutate(ticker = str_trim(as.character(ticker)),
         dlycaldt = ymd(dlycaldt)) %>%
  arrange(ticker, dlycaldt) %>%
  distinct(ticker, dlycaldt, .keep_all = TRUE)

# Derive quarter keys: current quarter and the prior completed quarter (avoids look-ahead bias)
crsp_merge_df <- crsp_merge_df %>%
  mutate(current_quarter_key = as.yearqtr(dlycaldt),
         prior_quarter_key = current_quarter_key - 0.25)

metric_feature_cols <- c(
  "net_positivity",
  "language_complexity",
  "numeric_transparency",
  "analyst_selectivity_ratio"
)

transcript_join_df <- transcripts_metric_df %>%
  rename(prior_quarter_key = quarter_key) %>%
  mutate(ticker = str_trim(as.character(ticker)))

merged_daily_df <- crsp_merge_df %>%
  left_join(transcript_join_df, by = c("ticker", "prior_quarter_key"))

# Transcript metrics arrive as strings from pivot; cast to numeric for modeling
for (col in metric_feature_cols) {
  merged_daily_df[[col]] <- suppressWarnings(as.numeric(merged_daily_df[[col]]))
}

rows_pre_md <- nrow(merged_daily_df)
merged_daily_df <- merged_daily_df %>%
  arrange(ticker, dlycaldt) %>%
  distinct(ticker, dlycaldt, .keep_all = TRUE)
cat("Post-merge dedup removed", rows_pre_md - nrow(merged_daily_df), "duplicate (ticker, dlycaldt) rows\n")

cat("crsp_merge_df shape :", paste(nrow(crsp_merge_df), ncol(crsp_merge_df), sep = ", "), "\n")
cat("merged_daily_df shape:", paste(nrow(merged_daily_df), ncol(merged_daily_df), sep = ", "), "\n")
head(merged_daily_df)
```

```{r cell_merged_info}
# --- Merged daily panel: dtypes and non-null counts ---
glimpse(merged_daily_df)
cat("\nNon-null counts:\n")
print(colSums(!is.na(merged_daily_df)))
```

```{r cell_next_day_return}
# 13) Primary model target: next_day_dlyretx (next row dlyretx). next_day_return is diagnostic (next row dlyret).
merged_daily_df <- merged_daily_df %>%
  arrange(ticker, dlycaldt) %>%
  group_by(ticker) %>%
  mutate(
    next_day_return = dplyr::lead(dlyret),
    next_day_dlyretx = dplyr::lead(dlyretx)
  ) %>%
  ungroup()

cat("next_day_return null count:", sum(is.na(merged_daily_df$next_day_return)), "\n")
cat("next_day_dlyretx null count:", sum(is.na(merged_daily_df$next_day_dlyretx)), "\n")
cat("Unique tickers (expect same NaN count for last-day rows):", length(unique(merged_daily_df$ticker)), "\n")
head(
  merged_daily_df %>%
    select(ticker, dlycaldt, dlyret, dlyretx, next_day_return, next_day_dlyretx),
  10
)
```

```{r cell_validation_checks}
# --- Data quality checks on merged panel (metrics vs returns) ---
# 14) Validation checks
metric_feature_cols <- c(
  "net_positivity",
  "language_complexity",
  "numeric_transparency",
  "analyst_selectivity_ratio"
)

# 1. Row preservation
crsp_row_count <- nrow(crsp_merge_df)
merged_row_count <- nrow(merged_daily_df)
cat(sprintf("Row preservation — CRSP: %s  |  Merged: %s\n",
            format(crsp_row_count, big.mark = ","),
            format(merged_row_count, big.mark = ",")))
stopifnot(crsp_row_count == merged_row_count)

# 2. Key uniqueness / duplicate-inflation check
crsp_dup_keys <- sum(duplicated(crsp_merge_df[, c("ticker", "dlycaldt")]))
dup_keys <- sum(duplicated(merged_daily_df[, c("ticker", "dlycaldt")]))
cat(sprintf("Duplicate (ticker, dlycaldt) in CRSP source : %d\n", crsp_dup_keys))
cat(sprintf("Duplicate (ticker, dlycaldt) after merge  : %d\n", dup_keys))

stopifnot(dup_keys == crsp_dup_keys)

# 3. Metric missingness after merge
missing_rates <- round(colMeans(is.na(merged_daily_df[, metric_feature_cols])) * 100, 2)
cat("\nMetric missing rate (%) after merge:\n")
print(missing_rates)

# 4–4b) next_day_return / next_day_dlyretx NaN audit — should only be NaN at last trading day per ticker

audit_one_target <- function(df, tgt_col, px_col) {
  n_t <- length(unique(df$ticker))
  last_day_per_ticker <- df %>%
    arrange(ticker, dlycaldt) %>%
    group_by(ticker) %>%
    summarise(dlycaldt = last(dlycaldt), .groups = "drop")
  last_day_join <- last_day_per_ticker %>%
    inner_join(df %>% select(ticker, dlycaldt, all_of(tgt_col)),
               by = c("ticker", "dlycaldt"))
  last_day_nulls <- sum(is.na(last_day_join[[tgt_col]]))
  non_last_nulls <- sum(is.na(df[[tgt_col]])) - last_day_nulls
  cat(sprintf("\n%s NaN at last date per ticker: %d / %d\n", tgt_col, last_day_nulls, n_t))
  cat(sprintf("%s NaN at non-last dates (unexpected): %d\n", tgt_col, non_last_nulls))

  evt <- df %>%
    arrange(ticker, dlycaldt) %>%
    group_by(ticker) %>%
    mutate(
      is_last_trade = dlycaldt == max(dlycaldt),
      px_lead = dplyr::lead(.data[[px_col]]),
      dlycaldt_lead = dplyr::lead(dlycaldt)
    ) %>%
    ungroup()

  anom <- evt %>% filter(is.na(.data[[tgt_col]]), !is_last_trade)
  cat(sprintf("\n=== %s anomaly audit (non-last NaN) ===\n", tgt_col))
  cat("Anomaly row count:", nrow(anom), "\n")
  if (nrow(anom) > 0) {
    show_cols <- c("ticker", "dlycaldt", px_col, "px_lead", tgt_col, "dlycaldt_lead")
    print(anom %>% select(any_of(show_cols)))
    cat("Interpretation: ", tgt_col, " should match ", px_col, " on next trade date; NaN usually means missing ",
        px_col, " on next trade date.\n", sep = "")
  }
  anom %>% distinct(ticker, dlycaldt)
}

drop_keys <- NULL
drop_keys <- bind_rows(
  audit_one_target(merged_daily_df, "next_day_return", "dlyret"),
  audit_one_target(merged_daily_df, "next_day_dlyretx", "dlyretx")
)

if (!is.null(drop_keys) && nrow(drop_keys) > 0) {
  cat("\nDropping these rows (reason code TARGET_NEXT_NA).\n")
  merged_daily_df <- merged_daily_df %>%
    anti_join(drop_keys, by = c("ticker", "dlycaldt")) %>%
    arrange(ticker, dlycaldt) %>%
    group_by(ticker) %>%
    mutate(
      next_day_return = dplyr::lead(dlyret),
      next_day_dlyretx = dplyr::lead(dlyretx)
    ) %>%
    ungroup()

  cat("\n--- After drop ---\n")
  audit_one_target(merged_daily_df, "next_day_return", "dlyret")
  audit_one_target(merged_daily_df, "next_day_dlyretx", "dlyretx")
}

# 5. Spot-check: quarter alignment for one ticker
sample_ticker <- merged_daily_df$ticker[1]
cat(sprintf("\nSpot-check for %s — date, current_quarter, prior_quarter, net_positivity:\n", sample_ticker))
print(head(merged_daily_df %>%
             filter(ticker == sample_ticker) %>%
             select(dlycaldt, current_quarter_key, prior_quarter_key, net_positivity), 10))
```

```{r cell_dedup}
# --- One row per (ticker, trade date) — idempotent (dedup already applied after merge) ---
merged_daily_df <- merged_daily_df %>%
  mutate(dlycaldt = ymd(as.character(dlycaldt)))

rows_before <- nrow(merged_daily_df)
merged_daily_df <- merged_daily_df %>%
  arrange(ticker, dlycaldt) %>%
  distinct(ticker, dlycaldt, .keep_all = TRUE)
rows_after <- nrow(merged_daily_df)

cat("Rows before:", rows_before, "\n")
cat("Rows after:", rows_after, "\n")
cat("Removed (expect 0 if upstream dedup succeeded):", rows_before - rows_after, "\n")
```

```{r cell_describe}
# --- Merged daily panel: numeric summary ---
summary(merged_daily_df)
```

## 8. Feature engineering and time-based split

- **Staged features (`FE_STAGE` 1–5)**: (1) baseline transcript levels + `dlyret` lags/rolls + market OHLCV; (2) add **`dlyretx` lags and rolling mean/std**; (3) add **OHLC microstructure** (overnight gap, intraday return, high–low range, Garman–Klass); (4) add **transcript dynamics** (QoQ deltas, deviation vs trailing 2-quarter mean, days since transcript quarter change); (5) add **cross-sectional ranks/z** by date and **industry-relative `dlyretx`** vs same-day `siccd` median.
- **Scaling**: **z-score** transcript level + dynamics columns with **train-only** means/SDs (non-finite values imputed at train means before scaling, matching `StandardScaler` behavior).
- **Governance**: duplicate-key assert and train/val calendar ordering check inside the feature chunk.
- **Split**: train / validation / test by calendar end dates (unchanged).

**Experimental protocol:** set `FE_STAGE` to `1` … `5` in the chunk below, then re-run the feature-engineering chunk, then the split + model chunks; compare RMSE / MAE / R² / directional accuracy / Spearman IC on the held-out test window. Default `FE_STAGE = 5` is the full pipeline.

```{r run_benchmark_config}
# Run A / Run B benchmark control
# A = strict no-transcript predictors
# B = full transcript pipeline
RUN_LABELS <- c("A", "B")
RUN_NAMES <- c(
  A = "Strict no-transcript",
  B = "Full transcript"
)
```

```{r fe_stage}
# Optional ablation control: set FE_STAGE to 1..5 before running feature engineering.
# 1 = baseline, 2 = +dlyretx lag/roll, 3 = +OHLC, 4 = +transcript dynamics, 5 = +cross-section (full).
FE_STAGE <- 5L
```

```{r cell_feature_engineering}
# ── Step 1: Feature engineering (mirrors Final_code.ipynb) ───────────────────
# Staged ablation: 1=baseline (levels+miss+market+dlyret lags/roll+cal+ids),
# 2=+dlyretx lag/rolling, 3=+OHLC microstructure, 4=+transcript dynamics, 5=+cross-section (full)
FE_STAGE <- if (exists("FE_STAGE")) as.integer(FE_STAGE[1]) else 5L
if (!FE_STAGE %in% 1L:5L) stop("FE_STAGE must be between 1 and 5")

model_df <- merged_daily_df %>%
  mutate(dlycaldt = ymd(as.character(dlycaldt))) %>%
  arrange(ticker, dlycaldt)

TRANSCRIPT_FEATS <- c(
  "net_positivity", "language_complexity",
  "numeric_transparency", "analyst_selectivity_ratio"
)

# Raw transcript snapshot (before scaling) for missingness + dynamics source
tr_raw <- model_df[, TRANSCRIPT_FEATS, drop = FALSE]

# ── Transcript dynamics (quarter-level) ───────────────────────────────────────
TRANSCRIPT_DYN_FEATS <- character(0)
if (FE_STAGE >= 4L && "prior_quarter_key" %in% names(model_df)) {
  dyn_base <- model_df %>%
    distinct(ticker, prior_quarter_key, .keep_all = TRUE) %>%
    arrange(ticker, prior_quarter_key)

  for (m in TRANSCRIPT_FEATS) {
    dyn_base <- dyn_base %>%
      group_by(ticker) %>%
      mutate(
        !!paste0(m, "_dq") := .data[[m]] - dplyr::lag(.data[[m]], 1L),
        !!paste0(m, "_dev2") := .data[[m]] - (dplyr::lag(.data[[m]], 1L) + dplyr::lag(.data[[m]], 2L)) / 2
      ) %>%
      ungroup()
  }
  dyn_join_cols <- c(
    "ticker", "prior_quarter_key",
    unlist(lapply(TRANSCRIPT_FEATS, function(m) c(paste0(m, "_dq"), paste0(m, "_dev2"))))
  )
  dyn_join_cols <- intersect(dyn_join_cols, names(dyn_base))
  model_df <- model_df %>%
    left_join(dyn_base %>% select(any_of(dyn_join_cols)), by = c("ticker", "prior_quarter_key"))

  model_df <- model_df %>%
    arrange(ticker, dlycaldt) %>%
    group_by(ticker) %>%
    mutate(
      q_prev = dplyr::lag(prior_quarter_key, 1L),
      ep_inc = dplyr::case_when(
        is.na(q_prev) ~ 1L,
        prior_quarter_key != q_prev ~ 1L,
        TRUE ~ 0L
      ),
      txn_ep = cumsum(ep_inc)
    ) %>%
    group_by(ticker, txn_ep) %>%
    mutate(days_since_transcript_change = as.integer(dlycaldt - min(dlycaldt))) %>%
    ungroup() %>%
    select(-q_prev, -ep_inc, -txn_ep)

  TRANSCRIPT_DYN_FEATS <- c(
    unlist(lapply(TRANSCRIPT_FEATS, function(m) c(paste0(m, "_dq"), paste0(m, "_dev2")))),
    "days_since_transcript_change"
  )
} else {
  for (m in TRANSCRIPT_FEATS) {
    model_df[[paste0(m, "_dq")]] <- NA_real_
    model_df[[paste0(m, "_dev2")]] <- NA_real_
  }
  model_df[["days_since_transcript_change"]] <- NA_real_
  TRANSCRIPT_DYN_FEATS <- c(
    unlist(lapply(TRANSCRIPT_FEATS, function(m) c(paste0(m, "_dq"), paste0(m, "_dev2")))),
    "days_since_transcript_change"
  )
}

# ── Lagged dlyret ─────────────────────────────────────────────────────────────
for (lag_k in c(1L, 2L, 5L, 10L, 20L)) {
  model_df <- model_df %>%
    group_by(ticker) %>%
    mutate(!!paste0("dlyret_lag", lag_k) := dplyr::lag(dlyret, lag_k)) %>%
    ungroup()
}

# ── dlyretx lag + rolling (FE_STAGE >= 2) ─────────────────────────────────────
RETX_LAG_FEATS <- paste0("dlyretx_lag", c(1, 2, 5, 10, 20))
RETX_ROLL_FEATS <- c(
  paste0("retx_mean_", c(5, 10, 20)),
  paste0("retx_std_", c(5, 10, 20))
)
if (FE_STAGE >= 2L) {
  for (lag_k in c(1L, 2L, 5L, 10L, 20L)) {
    model_df <- model_df %>%
      group_by(ticker) %>%
      mutate(!!paste0("dlyretx_lag", lag_k) := dplyr::lag(dlyretx, lag_k)) %>%
      ungroup()
  }
  for (w in c(5L, 10L, 20L)) {
    model_df <- model_df %>%
      group_by(ticker) %>%
      arrange(dlycaldt, .by_group = TRUE) %>%
      mutate(
        !!paste0("retx_mean_", w) := slider::slide_dbl(
          dplyr::lag(dlyretx, 1L), mean,
          .before = w - 1L, .complete = FALSE, na.rm = TRUE
        ),
        !!paste0("retx_std_", w) := slider::slide_dbl(
          dplyr::lag(dlyretx, 1L), stats::sd,
          .before = w - 1L, .complete = FALSE, na.rm = TRUE
        )
      ) %>%
      ungroup()
  }
} else {
  for (nm in RETX_LAG_FEATS) model_df[[nm]] <- NA_real_
  for (nm in RETX_ROLL_FEATS) model_df[[nm]] <- NA_real_
}

# ── Rolling stats on dlyret / dlyvol (shift by 1; windows 5 and 10) ───────────
for (window in c(5L, 10L)) {
  model_df <- model_df %>%
    group_by(ticker) %>%
    arrange(dlycaldt, .by_group = TRUE) %>%
    mutate(
      !!paste0("ret_mean_", window) := slider::slide_dbl(
        dplyr::lag(dlyret, 1L), mean,
        .before = window - 1L, .complete = FALSE, na.rm = TRUE
      ),
      !!paste0("ret_std_", window) := slider::slide_dbl(
        dplyr::lag(dlyret, 1L), stats::sd,
        .before = window - 1L, .complete = FALSE, na.rm = TRUE
      ),
      !!paste0("vol_mean_", window) := slider::slide_dbl(
        dplyr::lag(dlyvol, 1L), mean,
        .before = window - 1L, .complete = FALSE, na.rm = TRUE
      )
    ) %>%
    ungroup()
}

# ── OHLC microstructure ───────────────────────────────────────────────────────
MICROSTRUCT_FEATS <- c("overnight_gap", "intraday_return", "high_low_range", "gk_vol")
safe_div <- function(a, b, eps = 1e-12) {
  b2 <- ifelse(!is.na(b) & abs(b) < eps, sign(b) * eps, b)
  b2 <- ifelse(!is.na(b2) & b2 == 0, NA_real_, b2)
  ifelse(is.na(b2), NA_real_, a / b2)
}

model_df <- model_df %>%
  group_by(ticker) %>%
  mutate(dlyclose_lag1 = dplyr::lag(dlyclose, 1L)) %>%
  ungroup()

if (FE_STAGE >= 3L) {
  model_df <- model_df %>%
    mutate(
      overnight_gap = safe_div(dlyopen, dlyclose_lag1) - 1,
      intraday_return = safe_div(dlyclose, dlyopen) - 1,
      high_low_range = safe_div(dlyhigh - dlylow, ifelse(dlyclose == 0, NA_real_, dlyclose)),
      gk_vol = {
        lo <- log(safe_div(dlyhigh, dlylow))
        lc <- log(safe_div(dlyclose, dlyopen))
        0.5 * (lo^2) - (2 * log(2) - 1) * (lc^2)
      }
    )
} else {
  for (nm in MICROSTRUCT_FEATS) model_df[[nm]] <- NA_real_
}

# ── Cross-section + industry residual (FE_STAGE >= 5) ─────────────────────────
XSEC_FEATS <- character(0)
if (FE_STAGE >= 5L) {
  xsec_numeric <- c("dlyretx", "dlyret", "ret_mean_10", "ret_std_10", "vol_mean_10")
  for (xc in xsec_numeric) {
    if (!xc %in% names(model_df)) next
    model_df <- model_df %>%
      group_by(dlycaldt) %>%
      mutate(
        !!paste0(xc, "_cs_rank") := dplyr::percent_rank(.data[[xc]]),
        !!paste0(xc, "_cs_z") := {
          v <- .data[[xc]]
          m <- mean(v, na.rm = TRUE)
          s <- stats::sd(v, na.rm = TRUE)
          if (!is.finite(s) || s < 1e-12) rep(NA_real_, dplyr::n()) else (v - m) / s
        }
      ) %>%
      ungroup()
    XSEC_FEATS <- c(XSEC_FEATS, paste0(xc, "_cs_rank"), paste0(xc, "_cs_z"))
  }
  if ("siccd" %in% names(model_df)) {
    model_df <- model_df %>%
      group_by(dlycaldt, siccd) %>%
      mutate(dlyretx_ind_med = stats::median(dlyretx, na.rm = TRUE)) %>%
      ungroup() %>%
      mutate(dlyretx_ind_res = dlyretx - dlyretx_ind_med)
  } else {
    model_df$dlyretx_ind_med <- NA_real_
    model_df$dlyretx_ind_res <- NA_real_
  }
  XSEC_FEATS <- c(XSEC_FEATS, "dlyretx_ind_med", "dlyretx_ind_res")
} else {
  for (xc in c("dlyretx", "dlyret", "ret_mean_10", "ret_std_10", "vol_mean_10")) {
    model_df[[paste0(xc, "_cs_rank")]] <- NA_real_
    model_df[[paste0(xc, "_cs_z")]] <- NA_real_
    XSEC_FEATS <- c(XSEC_FEATS, paste0(xc, "_cs_rank"), paste0(xc, "_cs_z"))
  }
  model_df$dlyretx_ind_med <- NA_real_
  model_df$dlyretx_ind_res <- NA_real_
  XSEC_FEATS <- c(XSEC_FEATS, "dlyretx_ind_med", "dlyretx_ind_res")
}

# ── Calendar features (pandas dayofweek: Monday=0) ────────────────────────────
model_df <- model_df %>%
  mutate(
    month = lubridate::month(dlycaldt),
    dayofweek = (lubridate::wday(dlycaldt, week_start = 1) - 1L) %% 7L
  )

# ── Label-encode categorical identifiers (R tree APIs) ───────────────────────
for (col in c("ticker", "siccd", "naics")) {
  model_df[[paste0(col, "_enc")]] <- as.integer(factor(as.character(model_df[[col]]))) - 1L
}

# Missingness indicators from raw transcript NA
for (tf in TRANSCRIPT_FEATS) {
  model_df[[paste0("miss_", tf)]] <- as.integer(is.na(tr_raw[[tf]]))
}
model_df$miss_any_transcript <- as.integer(pmax(
  model_df$miss_net_positivity,
  model_df$miss_language_complexity,
  model_df$miss_numeric_transparency,
  model_df$miss_analyst_selectivity_ratio
))
TRANSCRIPT_MISS_FEATS <- c(paste0("miss_", TRANSCRIPT_FEATS), "miss_any_transcript")

# Train-only scaling on transcript levels + dynamics (when FE_STAGE >= 4)
TRAIN_END_SCALER <- as.Date("2021-12-31")
scaler_cols <- if (FE_STAGE >= 4L) {
  c(TRANSCRIPT_FEATS, intersect(TRANSCRIPT_DYN_FEATS, names(model_df)))
} else {
  TRANSCRIPT_FEATS
}

train_fit_mask <- model_df$dlycaldt <= TRAIN_END_SCALER
train_sub <- model_df[train_fit_mask, scaler_cols, drop = FALSE]
train_sub <- train_sub[stats::complete.cases(train_sub), , drop = FALSE]
if (nrow(train_sub) < 10) {
  scaler_cols <- TRANSCRIPT_FEATS
  train_sub <- model_df[train_fit_mask, scaler_cols, drop = FALSE]
  train_sub <- train_sub[stats::complete.cases(train_sub), , drop = FALSE]
}

mu_vec <- colMeans(train_sub, na.rm = TRUE)
sig_vec <- apply(train_sub, 2, stats::sd, na.rm = TRUE)
sig_vec[!is.finite(sig_vec) | sig_vec == 0] <- 1

for (nm in scaler_cols) {
  x <- as.numeric(model_df[[nm]])
  x[!is.finite(x)] <- mu_vec[[nm]]
  model_df[[nm]] <- (x - mu_vec[[nm]]) / sig_vec[[nm]]
}

MARKET_FEATS <- c("dlyret", "dlyretx", "dlyvol", "dlycap", "dlyclose", "dlyopen", "dlyhigh", "dlylow")
LAG_FEATS <- paste0("dlyret_lag", c(1, 2, 5, 10, 20))
ROLLING_FEATS <- c(
  paste0("ret_mean_", c(5, 10)),
  paste0("ret_std_", c(5, 10)),
  paste0("vol_mean_", c(5, 10))
)
CALENDAR_FEATS <- c("month", "dayofweek")
ID_FEATS <- c("ticker_enc", "siccd_enc", "naics_enc")

FEATURE_COLS <- c(
  TRANSCRIPT_FEATS, TRANSCRIPT_MISS_FEATS, MARKET_FEATS,
  LAG_FEATS, ROLLING_FEATS
)
if (FE_STAGE >= 2L) FEATURE_COLS <- c(FEATURE_COLS, RETX_LAG_FEATS, RETX_ROLL_FEATS)
if (FE_STAGE >= 3L) FEATURE_COLS <- c(FEATURE_COLS, MICROSTRUCT_FEATS)
if (FE_STAGE >= 4L) FEATURE_COLS <- c(FEATURE_COLS, intersect(TRANSCRIPT_DYN_FEATS, names(model_df)))
if (FE_STAGE >= 5L) FEATURE_COLS <- c(FEATURE_COLS, intersect(XSEC_FEATS, names(model_df)))
FEATURE_COLS <- c(FEATURE_COLS, CALENDAR_FEATS, ID_FEATS)
FEATURE_COLS <- intersect(FEATURE_COLS, names(model_df))

# ── Run A / Run B feature sets (strict benchmark) ─────────────────────────────
FEATURE_COLS_B <- FEATURE_COLS
dyn_cols_present <- intersect(TRANSCRIPT_DYN_FEATS, names(model_df))
TRANSCRIPT_EXCLUDE_COLS <- unique(c(
  TRANSCRIPT_FEATS,
  TRANSCRIPT_MISS_FEATS,
  dyn_cols_present
))
FEATURE_COLS_A <- setdiff(FEATURE_COLS_B, TRANSCRIPT_EXCLUDE_COLS)

RUN_FEATURE_SETS <- list(
  A = FEATURE_COLS_A,
  B = FEATURE_COLS_B
)
if (!exists("RUN_LABELS")) RUN_LABELS <- c("A", "B")
RUN_LABELS <- intersect(RUN_LABELS, names(RUN_FEATURE_SETS))
if (!length(RUN_LABELS)) stop("No valid RUN_LABELS found in RUN_FEATURE_SETS.")

# Backward-compatible default references use full model
FEATURE_COLS <- FEATURE_COLS_B

TARGET_COL <- "next_day_dlyretx"
model_df <- model_df %>% filter(!is.na(.data[[TARGET_COL]]))

stopifnot(!anyDuplicated(model_df[, c("ticker", "dlycaldt")]))
train_max <- max(model_df$dlycaldt[model_df$dlycaldt <= TRAIN_END_SCALER], na.rm = TRUE)
val_min <- min(model_df$dlycaldt[model_df$dlycaldt >= as.Date("2022-01-01")], na.rm = TRUE)
stopifnot(train_max < val_min)

cat(sprintf(
  "FE_STAGE=%d (1=baseline, 2=+dlyretx lag/roll, 3=+OHLC, 4=+transcript dyn, 5=+xsec/full)\n",
  FE_STAGE
))
cat(sprintf("model_df shape : %s\n", paste(nrow(model_df), ncol(model_df), sep = ", ")))
cat(sprintf("Features (%d): %s\n", length(FEATURE_COLS),
            paste0("['", paste(FEATURE_COLS, collapse = "', '"), "']")))
cat(sprintf("Target         : %s\n", TARGET_COL))
cat("\nTranscript missingness rate (% of days) — indicator means:\n")
print(round(colMeans(model_df[, TRANSCRIPT_MISS_FEATS, drop = FALSE]) * 100, 2))
dyn_cols <- intersect(TRANSCRIPT_DYN_FEATS, names(model_df))
cat("\nTranscript + dynamics — NaN % after scaling (expect ~0 on scaled cols):\n")
print(round(colMeans(is.na(model_df[, c(TRANSCRIPT_FEATS, dyn_cols), drop = FALSE])) * 100, 4))
cat("\nMissing rates (%) per feature:\n")
print(round(colMeans(is.na(model_df[, FEATURE_COLS, drop = FALSE])) * 100, 2))
cat("\nRun feature counts:\n")
for (run_id in RUN_LABELS) {
  cat(sprintf("  Run %s (%s): %d features\n",
              run_id,
              if (exists("RUN_NAMES") && run_id %in% names(RUN_NAMES)) RUN_NAMES[[run_id]] else run_id,
              length(RUN_FEATURE_SETS[[run_id]])))
}
strict_overlap <- intersect(RUN_FEATURE_SETS$A, TRANSCRIPT_EXCLUDE_COLS)
cat(sprintf("Run A transcript-overlap feature count (expect 0): %d\n", length(strict_overlap)))
stopifnot(length(strict_overlap) == 0L)
```

```{r cell_model_df_preview}
# --- Modeling matrix preview ---
head(model_df)
```

```{r cell_time_split}
# ── Step 2: Time-based train / validation / test split ───────────────────────
TRAIN_END  <- as.Date("2021-12-31")
VAL_START  <- as.Date("2022-01-01")
VAL_END    <- as.Date("2022-12-31")
TEST_START <- as.Date("2023-01-01")

train_df <- model_df %>% filter(dlycaldt <= TRAIN_END)
val_df   <- model_df %>% filter(dlycaldt >= VAL_START & dlycaldt <= VAL_END)
test_df  <- model_df %>% filter(dlycaldt >= TEST_START)

# Sanity checks — no overlap
stopifnot(max(train_df$dlycaldt) < min(val_df$dlycaldt))
stopifnot(max(val_df$dlycaldt)   < min(test_df$dlycaldt))

cat(sprintf("Train rows      : %10s  (%s → %s)   (%s)\n",
            format(nrow(train_df), big.mark = ","),
            as.character(min(train_df$dlycaldt)),
            as.character(max(train_df$dlycaldt)),
            paste(nrow(train_df), ncol(train_df), sep = ", ")))
cat(sprintf("Validation rows : %10s  (%s → %s)\n",
            format(nrow(val_df), big.mark = ","),
            as.character(min(val_df$dlycaldt)),
            as.character(max(val_df$dlycaldt))))
cat(sprintf("Test rows       : %10s  (%s → %s)\n",
            format(nrow(test_df), big.mark = ","),
            as.character(min(test_df$dlycaldt)),
            as.character(max(test_df$dlycaldt))))

# Build per-run matrices from the same split boundaries/rows.
RUN_SPLITS <- list()
for (run_id in RUN_LABELS) {
  run_cols <- RUN_FEATURE_SETS[[run_id]]
  X_train <- train_df[, run_cols, drop = FALSE]
  y_train <- train_df[[TARGET_COL]]
  X_val   <- val_df[, run_cols, drop = FALSE]
  y_val   <- val_df[[TARGET_COL]]
  X_test  <- test_df[, run_cols, drop = FALSE]
  y_test  <- test_df[[TARGET_COL]]
  RUN_SPLITS[[run_id]] <- list(
    feature_cols = run_cols,
    X_train = X_train, y_train = y_train,
    X_val = X_val, y_val = y_val,
    X_test = X_test, y_test = y_test
  )
}

# Row-level comparability guardrail across runs.
row_counts <- vapply(RUN_SPLITS, function(x) c(
  train = nrow(x$X_train),
  val = nrow(x$X_val),
  test = nrow(x$X_test)
), numeric(3L))
stopifnot(length(unique(row_counts["train", ])) == 1L)
stopifnot(length(unique(row_counts["val", ])) == 1L)
stopifnot(length(unique(row_counts["test", ])) == 1L)

# Backward-compatible aliases point to Run B (or first available run).
default_run <- if ("B" %in% names(RUN_SPLITS)) "B" else names(RUN_SPLITS)[1]
X_train <- RUN_SPLITS[[default_run]]$X_train
y_train <- RUN_SPLITS[[default_run]]$y_train
X_val   <- RUN_SPLITS[[default_run]]$X_val
y_val   <- RUN_SPLITS[[default_run]]$y_val
X_test  <- RUN_SPLITS[[default_run]]$X_test
y_test  <- RUN_SPLITS[[default_run]]$y_test
```

## 9. Model training, tuning, and comparison

**LightGBM** and **XGBoost** baselines, optional Random Forest, shared **evaluation** helpers, **CPU** setup in this R port (the Python notebook probes **GPU** where available), **walk-forward expanding-window time-series CV** on train+validation (by `dlycaldt`), refit best parameters on full train+val, a **comparison table**, and **feature importance** plots.

```{r cell_evaluate_helper}
# ── Step 3: Evaluation helper ─────────────────────────────────────────────────
evaluate <- function(y_true, y_pred, label = "") {
  # Return a list of regression and financial metrics; optionally print them.
  y_true <- as.numeric(y_true)
  y_pred <- as.numeric(y_pred)

  ok <- !is.na(y_true) & !is.na(y_pred)
  yt <- y_true[ok]
  yp <- y_pred[ok]

  rmse <- sqrt(mean((yt - yp)^2))
  mae  <- mean(abs(yt - yp))
  ss_res <- sum((yt - yp)^2)
  ss_tot <- sum((yt - mean(yt))^2)
  r2  <- 1 - ss_res / ss_tot
  dir_acc <- mean(sign(yt) == sign(yp))
  ic <- suppressWarnings(cor(yt, yp, method = "spearman"))

  result <- list(
    RMSE = rmse,
    MAE = mae,
    R2 = r2,
    Directional_Accuracy = dir_acc,
    Spearman_IC = ic
  )
  if (nzchar(label)) {
    cat(sprintf("\n── %s %s\n", label, strrep("─", max(0, 50 - nchar(label)))))
    for (k in names(result)) {
      cat(sprintf("  %-25s: %f\n", k, result[[k]]))
    }
  }
  result
}

# Container filled by model chunks below
results <- list()
models_by_run <- list()
tuning_by_run <- list()

cat("evaluate() helper registered. results dict initialised.\n")
```

```{r cell_lightgbm_baseline}
# ── Step 4a: LightGBM ────────────────────────────────────────────────────────
# Build lgb.Dataset with categorical_feature hint for label-encoded identifiers.
for (run_id in RUN_LABELS) {
  split_i <- RUN_SPLITS[[run_id]]
  X_train_mat <- data.matrix(split_i$X_train)
  X_val_mat   <- data.matrix(split_i$X_val)
  X_test_mat  <- data.matrix(split_i$X_test)
  cat_feats <- intersect(c("ticker_enc", "siccd_enc", "naics_enc"), colnames(split_i$X_train))

  dtrain_lgb <- lgb.Dataset(
    data = X_train_mat,
    label = split_i$y_train,
    categorical_feature = cat_feats
  )
  dval_lgb <- lgb.Dataset(
    data = X_val_mat,
    label = split_i$y_val,
    categorical_feature = cat_feats,
    reference = dtrain_lgb
  )

  lgb_params <- list(
    objective = "regression",
    learning_rate = 0.01,
    num_leaves = 127,
    min_data_in_leaf = 50,
    feature_fraction = 0.8,
    bagging_fraction = 0.8,
    max_depth = 6,
    verbose = -1
  )

  lgb_model <- lgb.train(
    params = lgb_params,
    data = dtrain_lgb,
    nrounds = 5000,
    valids = list(valid = dval_lgb),
    early_stopping_rounds = 100,
    eval = "l2"
  )

  lgb_preds <- predict(lgb_model, X_test_mat)
  key <- paste(run_id, "LightGBM", sep = "::")
  lbl <- sprintf("Run %s (%s) LightGBM — Test set (2023-2024)",
                 run_id,
                 if (exists("RUN_NAMES") && run_id %in% names(RUN_NAMES)) RUN_NAMES[[run_id]] else run_id)
  results[[key]] <- evaluate(split_i$y_test, lgb_preds, label = lbl)

  if (is.null(models_by_run[[run_id]])) models_by_run[[run_id]] <- list()
  models_by_run[[run_id]]$lgb_model <- lgb_model
  models_by_run[[run_id]]$cat_feats <- cat_feats
  models_by_run[[run_id]]$X_test_mat <- X_test_mat

  cat(sprintf("Run %s LightGBM best iteration: %s\n", run_id, lgb_model$best_iter))
}
```

```{r cell_xgboost_baseline}
# ── Step 4b: XGBoost ─────────────────────────────────────────────────────────
# CPU-only: cast the three *_enc columns to numeric, build xgb.DMatrix.
if (!exists("xgb_tree_method")) {
  xgb_tree_method <- "hist"
}

for (run_id in RUN_LABELS) {
  split_i <- RUN_SPLITS[[run_id]]
  cat_feats <- intersect(c("ticker_enc", "siccd_enc", "naics_enc"), colnames(split_i$X_train))

  X_train_xgb <- split_i$X_train
  X_val_xgb   <- split_i$X_val
  X_test_xgb  <- split_i$X_test
  for (col in cat_feats) {
    X_train_xgb[[col]] <- as.numeric(X_train_xgb[[col]])
    X_val_xgb[[col]]   <- as.numeric(X_val_xgb[[col]])
    X_test_xgb[[col]]  <- as.numeric(X_test_xgb[[col]])
  }

  dtrain_xgb <- xgb.DMatrix(data = data.matrix(X_train_xgb), label = split_i$y_train)
  dval_xgb   <- xgb.DMatrix(data = data.matrix(X_val_xgb),   label = split_i$y_val)
  dtest_xgb  <- xgb.DMatrix(data = data.matrix(X_test_xgb),  label = split_i$y_test)

  xgb_params <- list(
    objective = "reg:squarederror",
    tree_method = xgb_tree_method,  # set in GPU stub below to "hist"
    eta = 0.01,
    max_depth = 6,
    subsample = 0.8,
    colsample_bytree = 0.8
  )

  xgb_model <- xgb.train(
    params = xgb_params,
    data = dtrain_xgb,
    nrounds = 20000,
    watchlist = list(val = dval_xgb),
    early_stopping_rounds = 100,
    verbose = 0
  )

  xgb_preds <- predict(xgb_model, dtest_xgb)
  key <- paste(run_id, "XGBoost", sep = "::")
  lbl <- sprintf("Run %s (%s) XGBoost — Test set (2023-2024)",
                 run_id,
                 if (exists("RUN_NAMES") && run_id %in% names(RUN_NAMES)) RUN_NAMES[[run_id]] else run_id)
  results[[key]] <- evaluate(split_i$y_test, xgb_preds, label = lbl)

  if (is.null(models_by_run[[run_id]])) models_by_run[[run_id]] <- list()
  models_by_run[[run_id]]$xgb_model <- xgb_model
  models_by_run[[run_id]]$X_train_xgb <- X_train_xgb
  models_by_run[[run_id]]$X_val_xgb <- X_val_xgb
  models_by_run[[run_id]]$X_test_xgb <- X_test_xgb
  models_by_run[[run_id]]$dtest_xgb <- dtest_xgb

  cat(sprintf("Run %s XGBoost best iteration: %s\n", run_id, xgb_model$best_iteration))
}
```


```{r cell_tuning_setup}
# ── Tuning Setup: walk-forward time-series CV (mirrors Final_code.ipynb) ───────
# Expanding-window folds on `dlycaldt` over train+val only (pre-test).

TS_CV_N_FOLDS <- 4L
TS_CV_MIN_TRAIN_ROWS <- 5000L

array_split_vec <- function(x, k) {
  n <- length(x)
  if (k <= 0L) return(list())
  base <- n %/% k
  rem <- n %% k
  sizes <- rep(base, k)
  if (rem > 0L) sizes[seq_len(rem)] <- sizes[seq_len(rem)] + 1L
  starts <- c(1L, cumsum(utils::head(sizes, -1L)) + 1L)
  ends <- cumsum(sizes)
  lapply(seq_len(k), function(i) x[starts[i]:ends[i]])
}

make_expanding_date_folds <- function(d_trval, n_folds = TS_CV_N_FOLDS,
                                       min_train_rows = TS_CV_MIN_TRAIN_ROWS) {
  d <- as.Date(d_trval)
  u <- sort(unique(d))
  n_u <- length(u)
  n_folds <- as.integer(min(max(2L, n_folds), max(2L, n_u - 1L)))
  parts <- array_split_vec(u, n_folds + 1L)
  folds <- list()
  for (k in seq_len(n_folds)) {
    train_dates <- sort(unique(do.call(c, parts[seq_len(k)])))
    val_dates <- parts[[k + 1L]]
    if (!length(val_dates)) next
    tr_idx <- which(d %in% train_dates)
    va_idx <- which(d %in% val_dates)
    if (length(tr_idx) < min_train_rows || length(va_idx) == 0L) next
    stopifnot(max(d[tr_idx]) < min(d[va_idx]))
    folds[[length(folds) + 1L]] <- list(train = tr_idx, val = va_idx)
  }
  folds
}

summarize_ts_folds <- function(d_trval, folds, name) {
  d <- as.Date(d_trval)
  cat(sprintf("\n── %s: time-series folds (expanding by date) %s\n",
              name, strrep("─", max(0L, 12L))))
  for (i in seq_along(folds)) {
    tr <- folds[[i]]$train
    va <- folds[[i]]$val
    cat(sprintf(
      "  Fold %d: train %s rows  (%s → %s)  |  val %s rows  (%s → %s)\n",
      i,
      format(length(tr), big.mark = ","),
      as.character(min(d[tr])),
      as.character(max(d[tr])),
      format(length(va), big.mark = ","),
      as.character(min(d[va])),
      as.character(max(d[va]))
    ))
  }
}

run_context <- list()
for (run_id in RUN_LABELS) {
  split_i <- RUN_SPLITS[[run_id]]
  trval_df_i <- dplyr::bind_rows(train_df, val_df)
  d_trval_i <- trval_df_i$dlycaldt
  X_trval_i <- dplyr::bind_rows(split_i$X_train, split_i$X_val)
  y_trval_i <- c(split_i$y_train, split_i$y_val)
  stopifnot(max(as.Date(d_trval_i)) < min(test_df$dlycaldt))
  run_context[[run_id]] <- list(
    d_trval = d_trval_i,
    X_trval = X_trval_i,
    y_trval = y_trval_i,
    best_params_lgb = list(),
    best_params_xgb = list(),
    best_params_rf = list()
  )
  cat(sprintf("Run %s X_trval shape : %s  (train=%s + val=%s)\n",
              run_id,
              paste(nrow(X_trval_i), ncol(X_trval_i), sep = ", "),
              format(nrow(split_i$X_train), big.mark = ","),
              format(nrow(split_i$X_val),   big.mark = ",")))
}
cat("Tuning utilities ready (walk-forward CV by `dlycaldt`) for all runs.\n")
```

```{r cell_lgb_tuning}
# ── Tuning Cell B: LightGBM (time-series CV) ─────────────────────────────────
lgb_param_grid <- expand.grid(
  num_leaves        = c(63),   # c(63, 127, 255),
  min_child_samples = c(20),   # c(20, 100),
  learning_rate     = c(0.05), # c(0.01, 0.05),
  stringsAsFactors = FALSE
)

for (run_id in RUN_LABELS) {
  ctx <- run_context[[run_id]]
  folds_lgb <- make_expanding_date_folds(ctx$d_trval, TS_CV_N_FOLDS, TS_CV_MIN_TRAIN_ROWS)
  if (!length(folds_lgb)) stop(sprintf("No valid time-series folds for LightGBM (Run %s).", run_id))
  summarize_ts_folds(ctx$d_trval, folds_lgb, sprintf("LightGBM Run %s", run_id))

  cat(sprintf("\n── Run %s LightGBM: %d hyperparameter combos × %d folds %s\n",
              run_id, nrow(lgb_param_grid), length(folds_lgb), strrep("─", 8L)))

  best_mean <- Inf
  best_row <- NULL
  cat_feats <- intersect(c("ticker_enc", "siccd_enc", "naics_enc"), colnames(ctx$X_trval))
  for (j in seq_len(nrow(lgb_param_grid))) {
    params_i <- as.list(lgb_param_grid[j, , drop = FALSE])
    fold_rmses <- numeric(length(folds_lgb))
    for (fi in seq_along(folds_lgb)) {
      tr <- folds_lgb[[fi]]$train
      va <- folds_lgb[[fi]]$val
      dtrain_i <- lgb.Dataset(
        data = data.matrix(ctx$X_trval[tr, , drop = FALSE]),
        label = ctx$y_trval[tr],
        categorical_feature = cat_feats
      )
      dval_i <- lgb.Dataset(
        data = data.matrix(ctx$X_trval[va, , drop = FALSE]),
        label = ctx$y_trval[va],
        reference = dtrain_i
      )
      params_full <- list(
        objective = "regression",
        learning_rate = params_i$learning_rate,
        num_leaves = params_i$num_leaves,
        min_data_in_leaf = params_i$min_child_samples,
        feature_fraction = 0.8,
        bagging_fraction = 0.8,
        verbose = -1
      )
      model_i <- lgb.train(
        params = params_full,
        data = dtrain_i,
        nrounds = 1500,
        valids = list(valid = dval_i),
        early_stopping_rounds = 100,
        eval = "l2"
      )
      preds <- predict(model_i, data.matrix(ctx$X_trval[va, , drop = FALSE]))
      fold_rmses[fi] <- sqrt(mean((ctx$y_trval[va] - preds)^2))
    }
    mean_rmse <- mean(fold_rmses)
    cat(sprintf("  [Run %s %d/%d] mean_val_RMSE=%f  params=%s\n",
                run_id, j, nrow(lgb_param_grid), mean_rmse,
                paste(names(params_i), unlist(params_i), sep = "=", collapse = ", ")))
    if (mean_rmse < best_mean) {
      best_mean <- mean_rmse
      best_row <- params_i
    }
  }

  run_context[[run_id]]$best_params_lgb <- best_row
  if (length(run_context[[run_id]]$best_params_lgb) == 0L) {
    stop(sprintf("LightGBM tuning did not produce best_params_lgb for Run %s.", run_id))
  }
  cat(sprintf("\nRun %s best params (LightGBM): %s\n",
              run_id,
              paste(names(run_context[[run_id]]$best_params_lgb),
                    unlist(run_context[[run_id]]$best_params_lgb),
                    sep = "=", collapse = ", ")))
  cat(sprintf("Run %s best mean CV RMSE (across folds): %f\n", run_id, best_mean))
}
```

```{r cell_xgb_tuning}
# ── Tuning Cell C: XGBoost (time-series CV) ──────────────────────────────────
xgb_param_grid <- expand.grid(
  max_depth        = c(4, 6, 8),
  min_child_weight = c(1, 5),
  learning_rate    = c(0.01, 0.05),
  stringsAsFactors = FALSE
)

for (run_id in RUN_LABELS) {
  ctx <- run_context[[run_id]]
  cat_feats <- intersect(c("ticker_enc", "siccd_enc", "naics_enc"), colnames(ctx$X_trval))

  X_trval_xgb <- ctx$X_trval
  for (col in cat_feats) {
    X_trval_xgb[[col]] <- as.numeric(X_trval_xgb[[col]])
  }

  folds_xgb <- make_expanding_date_folds(ctx$d_trval, TS_CV_N_FOLDS, TS_CV_MIN_TRAIN_ROWS)
  if (!length(folds_xgb)) stop(sprintf("No valid time-series folds for XGBoost (Run %s).", run_id))
  summarize_ts_folds(ctx$d_trval, folds_xgb, sprintf("XGBoost Run %s", run_id))

  cat(sprintf("\n── Run %s XGBoost: %d hyperparameter combos × %d folds %s\n",
              run_id, nrow(xgb_param_grid), length(folds_xgb), strrep("─", 8L)))

  best_mean_x <- Inf
  best_row_x <- NULL
  for (j in seq_len(nrow(xgb_param_grid))) {
    params_i <- as.list(xgb_param_grid[j, , drop = FALSE])
    fold_rmses <- numeric(length(folds_xgb))
    for (fi in seq_along(folds_xgb)) {
      tr <- folds_xgb[[fi]]$train
      va <- folds_xgb[[fi]]$val
      dtrain_i <- xgb.DMatrix(
        data = data.matrix(X_trval_xgb[tr, , drop = FALSE]),
        label = ctx$y_trval[tr]
      )
      dval_i <- xgb.DMatrix(
        data = data.matrix(X_trval_xgb[va, , drop = FALSE]),
        label = ctx$y_trval[va]
      )
      params_full <- list(
        objective = "reg:squarederror",
        tree_method = xgb_tree_method,
        eta = params_i$learning_rate,
        max_depth = params_i$max_depth,
        min_child_weight = params_i$min_child_weight,
        subsample = 0.8,
        colsample_bytree = 0.8
      )
      model_i <- xgb.train(
        params = params_full,
        data = dtrain_i,
        nrounds = 15000,
        watchlist = list(val = dval_i),
        early_stopping_rounds = 100,
        verbose = 0
      )
      preds <- predict(model_i, dval_i)
      fold_rmses[fi] <- sqrt(mean((ctx$y_trval[va] - preds)^2))
    }
    mean_rmse <- mean(fold_rmses)
    cat(sprintf("  [Run %s %d/%d] mean_val_RMSE=%f  params=%s\n",
                run_id, j, nrow(xgb_param_grid), mean_rmse,
                paste(names(params_i), unlist(params_i), sep = "=", collapse = ", ")))
    if (mean_rmse < best_mean_x) {
      best_mean_x <- mean_rmse
      best_row_x <- params_i
    }
  }

  run_context[[run_id]]$best_params_xgb <- best_row_x
  if (length(run_context[[run_id]]$best_params_xgb) == 0L) {
    stop(sprintf("XGBoost tuning did not produce best_params_xgb for Run %s.", run_id))
  }
  cat(sprintf("\nRun %s best params (XGBoost): %s\n",
              run_id,
              paste(names(run_context[[run_id]]$best_params_xgb),
                    unlist(run_context[[run_id]]$best_params_xgb),
                    sep = "=", collapse = ", ")))
  cat(sprintf("Run %s best mean CV RMSE (across folds): %f\n", run_id, best_mean_x))
}
```



```{r cell_refit_and_compare}
# ── Tuning Cell E: Refit on train+val → evaluate on test ─────────────────────
for (run_id in RUN_LABELS) {
  split_i <- RUN_SPLITS[[run_id]]
  ctx <- run_context[[run_id]]
  cat_feats <- intersect(c("ticker_enc", "siccd_enc", "naics_enc"), colnames(ctx$X_trval))

  # ── 1. LightGBM tuned refit ─────────────────────────────────────────────────
  dtrval_lgb <- lgb.Dataset(
    data = data.matrix(ctx$X_trval),
    label = ctx$y_trval,
    categorical_feature = cat_feats
  )
  lgb_tuned_params <- list(
    objective = "regression",
    learning_rate    = ctx$best_params_lgb$learning_rate,
    num_leaves       = ctx$best_params_lgb$num_leaves,
    min_data_in_leaf = ctx$best_params_lgb$min_child_samples,
    feature_fraction = 0.8,
    bagging_fraction = 0.8,
    verbose = -1
  )
  lgb_tuned <- lgb.train(
    params = lgb_tuned_params,
    data = dtrval_lgb,
    nrounds = 1500
  )
  lgb_tuned_preds <- predict(lgb_tuned, data.matrix(split_i$X_test))
  results[[paste(run_id, "LightGBM_tuned", sep = "::")]] <- evaluate(
    split_i$y_test,
    lgb_tuned_preds,
    label = sprintf("Run %s (%s) LightGBM tuned — Test set (2023-2024)",
                    run_id,
                    if (exists("RUN_NAMES") && run_id %in% names(RUN_NAMES)) RUN_NAMES[[run_id]] else run_id)
  )

  # ── 2. XGBoost tuned refit ──────────────────────────────────────────────────
  X_trval_xgb <- ctx$X_trval
  X_test_xgb <- split_i$X_test
  for (col in cat_feats) {
    X_trval_xgb[[col]] <- as.numeric(X_trval_xgb[[col]])
    X_test_xgb[[col]] <- as.numeric(X_test_xgb[[col]])
  }
  dtrval_xgb <- xgb.DMatrix(data = data.matrix(X_trval_xgb), label = ctx$y_trval)
  dtest_xgb_refit <- xgb.DMatrix(data = data.matrix(X_test_xgb), label = split_i$y_test)
  xgb_tuned_params <- list(
    objective = "reg:squarederror",
    tree_method = xgb_tree_method,
    eta = ctx$best_params_xgb$learning_rate,
    max_depth = ctx$best_params_xgb$max_depth,
    min_child_weight = ctx$best_params_xgb$min_child_weight,
    subsample = 0.8,
    colsample_bytree = 0.8
  )
  xgb_tuned <- xgb.train(
    params = xgb_tuned_params,
    data = dtrval_xgb,
    nrounds = 1500,
    verbose = 0
  )
  xgb_tuned_preds <- predict(xgb_tuned, dtest_xgb_refit)
  results[[paste(run_id, "XGBoost_tuned", sep = "::")]] <- evaluate(
    split_i$y_test,
    xgb_tuned_preds,
    label = sprintf("Run %s (%s) XGBoost tuned — Test set (2023-2024)",
                    run_id,
                    if (exists("RUN_NAMES") && run_id %in% names(RUN_NAMES)) RUN_NAMES[[run_id]] else run_id)
  )

  if (is.null(models_by_run[[run_id]])) models_by_run[[run_id]] <- list()
  models_by_run[[run_id]]$lgb_tuned <- lgb_tuned
  models_by_run[[run_id]]$xgb_tuned <- xgb_tuned
}
# ── Summary comparison ────────────────────────────────────────────────────────
cat("\n\n══ Model comparison: baseline vs tuned ══\n")
compare_df <- dplyr::bind_rows(results, .id = "model")
model_parts <- do.call(rbind, strsplit(compare_df$model, "::", fixed = TRUE))
compare_df$run <- model_parts[, 1]
compare_df$model_name <- model_parts[, 2]
compare_df$run_name <- ifelse(compare_df$run %in% names(RUN_NAMES), RUN_NAMES[compare_df$run], compare_df$run)
compare_df <- compare_df[, c("run", "run_name", "model_name", "RMSE", "MAE", "R2", "Directional_Accuracy", "Spearman_IC")]
compare_df <- compare_df[order(compare_df$run, compare_df$model_name), ]
print(tibble::as_tibble(compare_df))

# Strict incremental comparison: Run B - Run A by model_name
metric_cols <- c("RMSE", "MAE", "R2", "Directional_Accuracy", "Spearman_IC")
compare_a <- compare_df[compare_df$run == "A", c("model_name", metric_cols), drop = FALSE]
compare_b <- compare_df[compare_df$run == "B", c("model_name", metric_cols), drop = FALSE]
delta_df <- merge(compare_b, compare_a, by = "model_name", suffixes = c("_B", "_A"), all = FALSE)
for (m in metric_cols) {
  delta_df[[paste0("Delta_", m, "_B_minus_A")]] <- delta_df[[paste0(m, "_B")]] - delta_df[[paste0(m, "_A")]]
}
cat("\n\n══ Incremental effect: Run B - Run A ══\n")
print(tibble::as_tibble(delta_df))
```

```{r cell_feature_importance}
# ── Step 5: Feature importance (gain) — baseline LightGBM / XGBoost ───────────
plot_importance_bars <- function(importances, feature_names, title, top_n = 20L) {
  importances <- as.numeric(importances)
  feature_names <- as.character(feature_names)
  ord <- order(importances)
  k <- min(top_n, length(importances))
  idx <- tail(ord, k)
  df_imp <- data.frame(
    feature = feature_names[idx],
    importance = importances[idx],
    stringsAsFactors = FALSE
  )
  ggplot(df_imp, aes(x = .data$importance, y = reorder(.data$feature, .data$importance))) +
    geom_col() +
    labs(title = title, x = "Importance", y = NULL)
}

for (run_id in RUN_LABELS) {
  run_name <- if (run_id %in% names(RUN_NAMES)) RUN_NAMES[[run_id]] else run_id
  if (!is.null(models_by_run[[run_id]]$lgb_model)) {
    lgb_imp_df <- lightgbm::lgb.importance(models_by_run[[run_id]]$lgb_model)
    print(plot_importance_bars(
      lgb_imp_df$Gain, lgb_imp_df$Feature,
      sprintf("Run %s (%s) LightGBM — Feature importance (gain); target: %s", run_id, run_name, TARGET_COL)
    ))
  }

  if (!is.null(models_by_run[[run_id]]$xgb_model)) {
    xgb_imp_df <- xgboost::xgb.importance(model = models_by_run[[run_id]]$xgb_model)
    print(plot_importance_bars(
      xgb_imp_df$Gain, xgb_imp_df$Feature,
      sprintf("Run %s (%s) XGBoost — Feature importance (total gain); target: %s", run_id, run_name, TARGET_COL)
    ))
  }
}

results_ic <- dplyr::bind_rows(results, .id = "model")
parts <- do.call(rbind, strsplit(results_ic$model, "::", fixed = TRUE))
results_ic$run <- parts[, 1]
results_ic$model_name <- parts[, 2]
results_ic <- results_ic[, c("run", "model_name", "RMSE", "MAE", "R2", "Directional_Accuracy", "Spearman_IC")]
results_ic <- results_ic[order(-results_ic$Spearman_IC), ]
cat("\n══ Best model by Spearman IC (test 2023–2024) ══\n")
cat(sprintf("Run %s | %s  (IC = %f)\n",
            results_ic$run[1], results_ic$model_name[1], results_ic$Spearman_IC[1]))
```
